Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-16 paper/section-general-framework.tex
New framework setup

\section{A General Framework}
\label{sec:a-general-framework}

%In this section 
We now present our framework for algebraically
indexed types. % and representation independence. %We define the syntax
%and semantics of algebraically indexed types, then their semantics
%and finally we present a general programming language for
%algebraically indexed types, together with an abstraction theorem.

@@

The index expressions and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We use the
  meta-syntactic variables $s,s_1,s_2,...$ for arbitrary sorts taken
  from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$. (We use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$.)
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$. %, describing the sorts
%  of the arguments of each primitive type.
\end{enumerate}

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups mentioned in \autoref{sec:motivating-examples}, so
  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
  %For the index operations, 
  We have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, \abs{\cdot} \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
  %  with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(\abs{\cdot}) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
%  The intended interpretations of the top three pairs of operations
%  are group unit, group combination and group negation,
%  respectively. 
%  When we discuss equational theories on index expressions 
%in \autoref{sec:type-equality} 
%we will impose the (abelian) group laws. For this example, 
We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
           [\SynGL{2}, \SynTransl{2}]$ and
           $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

We assume a countably infinite collection of index variable names $i,
i_1, i_2,\ldots$. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
..., i_n \mathord: s_n$ are lists of variable/sort pairs such that
all the variable names are distinct.
\begin{figure*}[t]
  \centering
  % \textbf{Index contexts}\\

@@

    \inferrule* [right=IOp]
    {\indexOp{f} \in \mathit{IndexOp} \\
      \indexOpArity(\indexOp{f}) = ([s_1,...,s_n], s) \\
      \{\Delta \vdash e_j : s_j\}_{1 \leq j \leq n}}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
  \end{mathpar}

@@

  \textbf{Well-indexed types}
  \begin{mathpar}
    \inferrule* [right=TyPrim]
    {\tyPrimNm{X} \in \mathit{PrimType} \\\\
      \primTyArity(\tyPrimNm{X}) = [s_1,...,s_n] \\
      \{\Delta \vdash e_j : s_j\}_{1\leq j \leq n}}
    {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \isType}

@@

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). 
We use $\tyPrimNm{bool}$ as an abbreviation for $\tyPrimNm{unit} + \tyPrimNm{unit}$.
The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use
We say that a type is \emph{quantifier-free} if it does not contain
ether universal or existential quantification.

@@

It %will be technically 
is convenient to express substitution of index
expressions %in our framework 
in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,...,e_n)$ such that $\Delta \vdash e_j :
s_j$ for all $1 \leq j \leq n$. Given a simultaneous substitution
$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
the set of all simultaneous substitutions $\sigma$ such that $\Delta
\vdash \sigma \Rightarrow \Delta'$.
%
%It will be useful to 
We can think of any sequence of sorts as an index
context. In particular, we will make use of simultaneous substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$. By
further abuse of notation, we write $\Delta \Rightarrow
\primTyArity(\tyPrimNm{X})$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$.

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. %Note that we have implicity used the fact that
%well-sortedness of index expressions is preserved by addition of extra
%items to the context.
%
Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  %Similarly, 
Given
%a well-indexed type 
$\Delta' \vdash A \isType$, 
%can apply $\sigma$ to $A$ to produce a new well-indexed type 
we have $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}
The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
%$\Delta$: 
$\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of 
%two simultaneous substitutions 
$\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
%
Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

@@

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $\abs{\cdot}$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

The equality judgment $\Delta \vdash e \equiv e' : s$ on index
expressions generates the equality judgment $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating equality judgments on
types equates %states that %two 
applications of primitive types %are equal 
if
their %index expression 
arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for equality on types ensure that it is a
congruence relation 
and an equivalence relation.
%on types. %and that it is an equivalence relation
%(i.e.,~equality on types is reflexive, symmetric and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics of Algebraically-Indexed Types}
\label{sec:semantics-algebraically-indexed-types}

%\fixme{Make sure that the semantics is motivated}
%Having defined the language of algebraically-indexed types, we now
%turn to their denotational interpretation. 
%We now give the denotational interpretation of algebraically indexed
%types.  In \autoref{sec:index-erasure-semantics} we define an
%\emph{index-erasure} interpretation of types that interprets every
%well-indexed type as a set, ignoring the indexing
%expressions. %Building on the index-erasure semantics, we 
%We then define in
%\autoref{sec:relational-semantics} the relational
%interpretation of types.

\autoref{sec:index-erasure-semantics} gives an \emph{index-erasure}
interpretation of types, interpreting every well-indexed type as a
set, ignoring the indexing expressions.
\autoref{sec:relational-semantics} gives the relational interpretation
of types.

\subsubsection{The Index-Erasure Interpretation of Types}
\label{sec:index-erasure-semantics}

The defining feature of the index erasure interpretation is that 
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. %Accordingly, 
We thus assume each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$ and %We 
extend this assignment to %every
well-indexed types by: % induction on the type structure:
\begin{displaymath}
  \begin{array}{@{}c@{\hspace{0em}}c@{}}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
% The interpretation of the unit type is a chosen one element set, and
% the interpretations of the function, tuple and sum types are simply
% the corresponding constructions on sets. 
Note that 
the interpretations of the
universal and existential quantifiers 
do indeed
ignore the indexing. %: the
%interpretation of the type $\forall i\mathord:s.A$ is exactly the
%interpretation of the type $A$, and likewise for $\exists
%i\mathord:s.A$.
We then have:

\begin{example}[Two-Dimensional Geometry]
  \exref{ex:two-dim-geo-axioms}
%  The two-dimensional geometry instantiation of the general framework
  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
\end{example}

\subsubsection{The Relational Interpretation of Types}
\label{sec:relational-semantics}

For the unit, tuple, sum and function types we define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types require an interpretation of index
contexts. %We assign interpretations to index contexts 
These are given in terms of sets
of \emph{relational environments}.

The definition of relational environments is the most subtle part of
our framework.

>>>>

\newcommand{\Gen}{\mathrm{Gen}}
\newcommand{\Free}{\mathrm{Free}}
\newcommand{\semSort}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
\newcommand{\semIndexExp}[2]{\llbracket #1 \rrbracket^{\mathcal{#2}}}
\newcommand{\semPrim}[3]{\llbracket #1 \rrbacket^{\mathcal{#2}}_{\mathcal{#3}}}
\newcommand{\opsCtxt}{\Gamma_{\mathit{ops}}}
\newcommand{\opsEnv}{\eta_{\mathit{ops}}}
\section{A General Framework}
\label{sec:a-general-framework}

%In this section 
We now present our framework for algebraically indexed types and its
relational interpretation.  We define the syntax of algebraically
indexed types (\autoref{sec:algebraically-indexed-types}) and a syntax
for terms in a general programming language for algebraically indexed
types (\autoref{sec:well-typed-programs}). We give an index-erasure
semantics to types and terms (\autoref{sec:erasure-semantics}), and
based on this semantics define a notion of contextual equivalence. We
then introduce a relational semantics for types parameterised by an appropriate
`model' of the algebraic theory (\autoref{sec:relational-semantics}),
prove the central Abstraction theorem
(\autoref{sec:abstraction-theorem}), and use the relational semantics
to define a notion of semantic equivalence which we prove is a sound
approximation of contextual equivalence.

%and semantics of algebraically indexed types, then their semantics
%and finally we present a general programming language for
%algebraically indexed types, together with an abstraction theorem.

@@

The index expressions and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We use the
  meta-syntactic variables $s,s_1,s_2,\ldots$ for arbitrary sorts taken
  from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$. (We use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$.)
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$. %, describing the sorts
%  of the arguments of each primitive type.
\end{enumerate}

\begin{example*}[Geometry: syntax]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups mentioned in \autoref{sec:motivating-examples}, so
  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
  %For the index operations, 
  We have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, \abs{\cdot} \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
  %  with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(\abs{\cdot}) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
%  The intended interpretations of the top three pairs of operations
%  are group unit, group combination and group negation,
%  respectively. 
%  When we discuss equational theories on index expressions 
%in \autoref{sec:type-equality} 
%we will impose the (abelian) group laws. For this example, 
We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
           [\SynGL{2}, \SynTransl{2}]$ and
           $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\qed
\end{example*}

We assume a countably infinite collection of index variable names $i,
i_1, i_2$, \emph{etc}. \emph{Index contexts} $\Delta = i_1 \mathord: s_1,
\ldots, i_n \mathord: s_n$ are lists of variable/sort pairs such that
all the variable names are distinct.
\begin{figure*}[t]
  \centering
  % \textbf{Index contexts}\\

@@

    \inferrule* [right=IOp]
    {\indexOp{f} \in \mathit{IndexOp} \\
      \indexOpArity(\indexOp{f}) = ([s_1,\ldots,s_n], s) \\
      \{\Delta \vdash e_j : s_j\}_{1 \leq j \leq n}}
    {\Delta \vdash \indexOp{f}(e_1, \ldots, e_n) : s}
  \end{mathpar}

@@

  \textbf{Well-indexed types}
  \begin{mathpar}
    \inferrule* [right=TyPrim]
    {\tyPrimNm{X} \in \mathit{PrimType} \\\\
      \primTyArity(\tyPrimNm{X}) = [s_1,\ldots,s_n] \\
      \{\Delta \vdash e_j : s_j\}_{1\leq j \leq n}}
    {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \isType}

@@

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). 
We use $\tyPrimNm{bool}$ as an abbreviation for $\tyPrimNm{unit} + \tyPrimNm{unit}$.
The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,\ldots,e_n$, the well-indexed type
$\tyPrim{X}{e_1,\ldots,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use

@@

It %will be technically 
is convenient to express substitution of index
expressions %in our framework 
in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, \ldots, i_n \mathord: s_n$, a \emph{simultaneous substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,\ldots,e_n)$ such that $\Delta \vdash e_j :
s_j$ for all $1 \leq j \leq n$. Given a simultaneous substitution
$\Delta \vdash \sigma = (e_1,\ldots,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$. We write $\Delta \Rightarrow \Delta'$ for
the set of all simultaneous substitutions $\sigma$ such that $\Delta
\vdash \sigma \Rightarrow \Delta'$.
%
%It will be useful to 
We can think of any sequence of sorts as an index
context. In particular, we will make use of simultaneous substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$. By
further abuse of notation, we write $\Delta \Rightarrow
\primTyArity(\tyPrimNm{X})$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$.

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), \ldots, \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. %Note that we have implicity used the fact that
%well-sortedness of index expressions is preserved by addition of extra
%items to the context.
%
Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,\ldots,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, \ldots, \sigma^*e_n)$.  %Similarly, 
Given
%a well-indexed type 
$\Delta' \vdash A \isType$, 
%can apply $\sigma$ to $A$ to produce a new well-indexed type 
we have $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,\ldots,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,\ldots,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}
The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
%$\Delta$: 
$\id_\Delta = (i_1,\ldots,i_n)$ where $\Delta =
i_1\mathord:s_1,\ldots,i_n\mathord:s_n$. The \emph{composition} of 
%two simultaneous substitutions 
$\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,\ldots,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, \ldots, \sigma^*e'_n) \Rightarrow \Delta''$.
%
Given a context $\Delta = i_1\mathord:s_1,\ldots,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,\ldots,i_n)$. %The subscript on
%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

@@

\begin{example*}[Geometry: axioms]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $\abs{\cdot}$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\qed
\end{example*}

The equality judgment $\Delta \vdash e \equiv e' : s$ on index
expressions generates the equality judgment $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating equality judgments on
types equates %states that %two 
applications of primitive types %are equal 
if
their %index expression 
arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,\ldots,e_n} \equiv \tyPrim{X}{e'_1,\ldots,e'_n} \isType}
\end{displaymath}
The rest of the rules for equality on types ensure that it is a
congruence relation 
and an equivalence relation.
%on types. %and that it is an equivalence relation
%(i.e.,~equality on types is reflexive, symmetric and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Well-typed Programs}
\label{sec:well-typed-programs}

We now present the rules for well-typed programs over the 
collection of 
types we defined 
in \autoref{sec:algebraically-indexed-types}.

%Each
%well-typed program is assigned an index-erasure semantics, building on
%the index-erasure semantics of types %we defined 
%in \autoref{sec:index-erasure-semantics}. Our main result
%(\thmref{thm:abstraction}) is that the index-erasure semantics of
%every well-typed program is related to itself in the relational
%interpretation of its type: this is the abstraction theorem for every
%instantiation of our general framework.
